Data Preparation

### 2019-01-10 Data Preparation.Rmd
debugging course data

# Variable Engineering
##Course Data
### AoD
In this analysis, we conside that a course can cover an AoD in two ways: a course covers an AoD if one of its ILOs covers it or if one of its assessments cover it. For instance, if one of the ILO of a course states that the students will learn to analyze empirical data in the context of academic research, then the course in question covers the AoD `Research Skills`; and if one of the assessment is a group presentation, then the course also covers the AoD `Collaborative Skills` and `Communication Skills`.

#### AoD from ILOs
In order to determine which AoD each course covers with its ILOs, we first eliminate the AoD that are not covered by the ILOs and then we take the unique combination of course and AoD in case a course had several ILOs covering the same AoD.
```{r d_AoD_ILO}
d_AoD_ILO <- d_ILO %>%
  select(Code, AoD) %>%
  unique

@@

#### AoD from Assessment

In order to determine which AoD each course covers with its assessments, we need to create a binary matrix which indicates which AoD(s) each assessment type covers. We have created such matrix on our google drive and we save it in the following piece of code as `map_assessment_AoD`. `map_assessment_AoD` indicates that, for instance, the assessment type `Essay` covers the AoD `Critical Thinking Skills`, `Communication Skills` and `Writing Skill`.
```{r map_assessment_AoD}
map_assessment_AoD <- as.matrix(gsheet2tbl('https://docs.google.com/spreadsheets/d/1soRA1u5zf9oLNirGmZ9yZ7m2ccPa3XFemnxkj5AVRXo/edit#gid=719531216')[,lists$AoD] %>%
                                  mutate_all(as.logical))
rownames(map_assessment_AoD) <- lists$Assessment
print(map_assessment_AoD[1:5, 8:12])
```

@@

#### Combining AoD from ILOs and from assessments
Finally, we can use a `rbind` to combine the two tibbles `d_AoD_ILO` and `d_AoD_assessment`. We also use `unique` in case a course covers an AoD with both its ILOS and its assessments.
```{r rbind d_AoD_ILO and d_AoD_assessment}
d_AoD <- rbind(d_AoD_ILO,
               d_AoD_assessment) %>%
  distinct %>%
  arrange(Code)

@@

### Courses
Now that we have a clear overview of the distribution of AoDs (`d_AoD`) and assessments (`d_assessment`) among the courses, let us add variables to the tibble `d_course` that contain the information at the course level. The tibble `d_ILO_detail` contains two columns indicating the code of the course and the number of ILOs it contains; `d_assessment_detail` contains three columns indicating the code of the course, the number of assessments it covers and a list of the assessments it covers; `d_AoD_detail` contains three columns indicating the code of the course, the number of AoD it covers and a list of the AoD it covers. We use a `full_join` to add these variables to the tibble `d_course`.
```{r}
d_ILO_detail <- d_ILO %>% 
  select(Code, ILO) %>%
  unique %>%
  count(Code) %>%
  rename(n_ILO = n)
print(d_ILO_detail)

d_assessment_detail <- d_assessment %>%
  group_by(Code) %>%
  summarise(n_assessment = n(),
            `Assessments Covered` = paste(Assessment, collapse = ", "))
print(d_assessment_detail)

d_AoD_detail <- d_AoD %>%
  group_by(Code) %>%
  summarise(n_AoD = n(),
            `AoD Covered` = paste(AoD, collapse = ", "))
print(d_AoD_detail)
```

```{r}
d_course <- d_course %>%
  full_join(d_ILO_detail       , by = "Code") %>%
  full_join(d_assessment_detail, by = "Code") %>%
  full_join(d_AoD_detail       , by = "Code") %>%
  select(Code, `Course Title`, Cluster, n_ILO, n_assessment, `Assessments Covered`, n_AoD, `AoD Covered`, everything())
print(d_course)
```

The undegraduate research courses (*UGR-*) are only present at the `3000` level (advanced level). Yet, these course are also offered at the `2000` level (intermediate level). We use an `rbind` to duplicate the rows of the course `UGR3000` and mutate their `Code` to `UGR2000`.
```{r}
d_course <- d_course %>%
  rbind(filter(., Code %in% c("UGR3001", "UGR3002", "UGR3003", "UGR3005")) %>%
          mutate(Code = case_when(Code == "UGR3001" ~ "UGR2001",
                                  Code == "UGR3002" ~ "UGR2002",
                                  Code == "UGR3003" ~ "UGR2003",
                                  Code == "UGR3005" ~ "UGR2005")))

@@

Finally, we add a series of informative variable at the course level.
```{r}
d_course <- d_course %>%
  mutate(
    `Missing from ILO File` = as.logical(`Missing from ILO File`),
    Letters = substring(Code, 1, 3),
    Number  = as.numeric(substring(Code, 4, 7)),
    Level   = case_when(between(Number, 1000, 1999) ~ "Introductory",
                        between(Number, 2000, 2999) ~ "Intermediate",
                        between(Number, 3000, 3999) ~ "Advanced"    ))
print(select(d_course, Code, Letters, Number, Level))
```

##Textual Data
### Course Catalogue
In order to conduct a topic analysis of course content, we extract the *overview* and *description* of each course from the course catalogues.

#### Extracting Overviews
We start by extracting the overviews which are 1 or 2 page long and contain important information of a course. To accomplish this, for each catalogue (loop), we first use `grep` to determine which pages should be excluded from the analysis (`pages_to_exclude`). These pages are typically headers. Excluding them before the analysis allows use to keep the code relatively simple. We also make use the fact that the overviews always start with the code of the course, i.e. it starts with one of the elements of `course_code` and use `grep` to identify the first page of each overview. We use grep a 3rd time to identify the page containing the header *Core Coureses (COR)* and which marks the beginning of the overviews in the catalogue, and we loop from this page to the last page of the catalogue.

@@

#### Extracting Descriptions
Since the overviews contain a lot of information that does not interest us, we extract the description section of the overviews. The description section contains a brief description (200-500 words) of the content of the course. To accomplish this, for each overview, we need to identify the line where the description starts and the line where it ends. The description section is always preceeded by a header saying *Description* or *Course Description* (`start_descrip`), making it relatively easy to find the starting line of the decription section with `grep`. The description section is usually followed by the literature section which starts with header saying *Literature*, *Recommended Literature* or similar (see `end_descrip`). Using `grep` together with `end_descrip`, we can again identify the ending line of the description section of almost overviews. A handful overviews do not contain a *Literature* section. For these, the section *Instructional Format* marks the end of the description section.
```{r description}
# Setup
symbol_to_line <- c("\r\n", "\n")
symbol_to_line <- paste(symbol_to_line, collapse = "|")

@@

#### Tidy Text Format
We first create two tibbles `d_overview_tidy` and `d_description_tidy` that respectively store the overviews and the descriptions in the tidy text format, with one row per course-year-word.
```{r tidy text data}
d_overview_tidy <- d_description %>%
  select(Code, `Calendar Year`, Overview) %>%
  unnest_tokens(output = word, input = Overview)
print(d_overview_tidy)

@@

#### Stemming
We then use the function `hunspell__stem`, which returns valid stems for a given word, to stem the words in the tibbles `d_overview_tidy` and `d_description_tidy`. We first create a function `stem_hunspell` which, given a term, returns its most basic stem (the last one from the list of stem returned by `hunspell__stem`). We then apply `stem_hunspell` on the words of `d_overview_tidy`. Since, `stem_hunspell` is not a vectorized function and the number of words in `d_overview_tidy` is large, we first use distinct on the `word` variable to find the a list containing *once* each term present in the course overviews and apply the function `stem_hunspell` on this list. We save the terms and their stems in the tibble `stems`. Since the course descriptions can be conceptualized as a subset of the course overviews, the tibble `stems` contains all the terms present in the descriptions and can therefore safely used to stem the words in `d_description_tidy`.
```{r function stem_hunspell, cache = TRUE}
stem_hunspell <- function(term) {
  # look up the term in the dictionary
  stems <- hunspell_stem(term)[[1]]

@@

### Course Manual 
We create a dataframe from the Corpus:
```{r}
manuals_td <- tidy(corpus_manuals)
```

@@

```{r}
manuals_tokens <- manuals_td %>%
  select(id,text) %>%
  mutate(Concentration= substring(id, 1, 3), Code= substring(id,1,7))%>%
  unnest_tokens(word, text)
```
####(Optional) Pre-processing Course Manuals- Removing stopwords
Some characters appear with sufficient frequency to be relevant but not often enough accross all courses to be deleted when using tf-idf (see sections bellow), it is therefore convenient to add some stopwords and remove numbers. For this we import stop_words from tidytext, and search all our words with grep for stand alone numbers. We save these words in different dataframes and then exclude with an antijoin.
```{r}

@@

##Student Data
###Renaming Variables
First we rename the variable names to make them more comprehensible:
```{r}
d_rename <-d_dirt %>%
  rename(
    ID                     =  "X...Student.Number", 
    Code                   =  "Module..Abbrev..", 
    `Course Title`         =  "Module..Desc..",
    `Program Abbreviation` =  "Program..Abbreviation.", 
    `Program Description`  =  "Program..Description.",               
    Year                   =  "Academic.Year",                      
    Session                =  "Academic.Session",
    `Appraisal Code`       =  "Appraisal.Type",                     
    `Appraisal Title`      =  "Appraisal..Description.",             
    Attempt                =  "Number.rep..attemp",                 
    Grade                  =  "Grade.symbol",                        
    `Appraisal Status`     =  "Appraisal.Status",                   
    `Appraisal Status Description` =  "Appraisal.Status..Desc..",    #is only NA: unique(d4$Appraisal.Status..Desc..)       
    `Booking Reason`       =  "Module.Booking.Reason..Description.",
    `Appraisal Date`       =  "Appraisal.date",                      
    `Attempted Credits`    =  "Attempted.credits",                  
    `Earned Credit`        =  "Earned.Credits",                      
    `Credit Type`          =  "Credit.Type",                        
    `Personnel Number`     =  "Personnel.Number",                  
    `Name (Employee/Applicant)`=  "Name.of.employee.or.applicant",      
    `Individual Work`      =  "Individual.Work",                     
    `Academic Work ID`     =  "Academic.Work.ID",                   
    `Booking Status`       =  "Booking.Status",                      #empty
    `Booking Status Description` =  "Booking.Status..Desc..",             
    `Event Object Id`      =  "Event.Object.Id",                     #empty forwards and including      
    `Business Event Abbreviation` =  "Business.event.abbr.",               
    `Event Description`    =  "Event.Description",                   
    `Start time`           =  "Start.time",                         
    `End time`             =  "End.time",                            
    `Date`                 =  "Date",                               
    `Studyplan`            =  "Studyplan",                           
    `Conditional booking`  =  "Conditiona.booking",                 
    `Wachtlist`            =  "Wachtlist",                           
    `Object ID`            =  "Object.ID",                          
    `Object Name`          =  "Object.name",                         
    `Start Date`           =  "Start.date",                         
    `End Date`             =  "End.Date",                            
    `Number of Weeks`      =  "Number.of.weeks",                      #until this one next has values              
    `Appraiser ID`         =  "Appraiser.ID",                        
    `Appraiser Name`       =  "Appraiser.Name"

>>>>

# Variable Engineering

## AoD
In this analysis, we conside that a course can cover an AoD in two ways: a course covers an AoD if one of its ILOs covers it or if one of its assessments cover it. For instance, if one of the ILO of a course states that the students will learn to analyze empirical data in the context of academic research, then the course in question covers the AoD `Research Skills`; and if one of the assessment is a group presentation, then the course also covers the AoD `Collaborative Skills` and `Communication Skills`.

### AoD from ILOs
In order to determine which AoD each course covers with its ILOs, we first eliminate the AoD that are not covered by the ILOs and then we take the unique combination of course and AoD in case a course had several ILOs covering the same AoD.
```{r d_AoD_ILO}
d_AoD_ILO <- d_ILO %>%
  select(Code, AoD) %>%
  unique

@@

### AoD from Assessments
In order to determine which AoD each course covers with its assessments, we need to create a binary matrix which indicates which AoD(s) each assessment type covers. We have created such matrix on our google drive and we save it in the following piece of code as `map_assessment_AoD`. `map_assessment_AoD` indicates that, for instance, the assessment type `Essay` covers the AoD `Critical Thinking Skills`, `Communication Skills` and `Writing Skill`.
```{r map_assessment_AoD}
map_assessment_AoD <- as.matrix(gsheet2tbl('https://docs.google.com/spreadsheets/d/1soRA1u5zf9oLNirGmZ9yZ7m2ccPa3XFemnxkj5AVRXo/edit#gid=719531216')[,lists$AoD] %>%
                                  mutate_all(as.logical))
rownames(map_assessment_AoD) <- lists$Assessment
print(map_assessment_AoD[1:5, 8:12])
```

@@

### Combining AoD from ILOs and from assessments
Finally, we can use a `rbind` to combine the two tibbles `d_AoD_ILO` and `d_AoD_assessment`. We also use `unique` in case a course covers an AoD with both its ILOS and its assessments.
```{r rbind d_AoD_ILO and d_AoD_assessment}
d_AoD <- rbind(d_AoD_ILO,
               d_AoD_assessment) %>%
  distinct %>%
  arrange(Code)

@@

## Course Data
Now that we have a clear overview of the distribution of AoDs (`d_AoD`) and assessments (`d_assessment`) among the courses, let us add variables to the tibble `d_course` that contain the information at the course level. The tibble `d_ILO_detail` contains two columns indicating the code of the course and the number of ILOs it contains; `d_assessment_detail` contains three columns indicating the code of the course, the number of assessments it covers and a list of the assessments it covers; `d_AoD_detail` contains three columns indicating the code of the course, the number of AoD it covers and a list of the AoD it covers. We use a `full_join` to add these variables to the tibble `d_course`.
```{r d_course detail}
d_ILO_detail <- d_ILO %>% 
  select(Code, ILO) %>%
  distinct() %>%
  count(Code) %>%
  rename(n_ILO = n)

d_assessment_detail <- d_assessment %>%
  group_by(Code) %>%
  summarise(n_assessment = n(),
            `Assessments Covered` = paste(Assessment, collapse = ", "))

d_AoD_detail <- d_AoD %>%
  group_by(Code) %>%
  summarise(n_AoD = n(),
            `AoD Covered` = paste(AoD, collapse = ", "))

d_course_detail <- d_ILO_detail %>%
  full_join(d_assessment_detail, by = "Code") %>%
  full_join(d_AoD_detail       , by = "Code")

rm(d_ILO_detail, d_assessment_detail, d_AoD_detail)
print(d_course_detail)
```

```{r join d_course_detail}
d_course <- full_join(d_course, d_course_detail, by = "Code") %>%
  select(Code, `Course Title`, Cluster, n_ILO, n_assessment, `Assessments Covered`, n_AoD, `AoD Covered`, everything())

rm(d_course_detail)
print(d_course)
```

The undegraduate research courses (*UGR-*) are only present at the `3000` level (advanced level). Yet, these course are also offered at the `2000` level (intermediate level). We use an `rbind` to duplicate the rows of the course `UGR3000` and mutate their `Code` to `UGR2000`.
```{r create UGR2000}
d_course <- d_course %>%
  rbind(filter(., Code %in% c("UGR3001", "UGR3002", "UGR3003", "UGR3005")) %>%
          mutate(Code = case_when(Code == "UGR3001" ~ "UGR2001",
                                  Code == "UGR3002" ~ "UGR2002",
                                  Code == "UGR3003" ~ "UGR2003",
                                  Code == "UGR3005" ~ "UGR2005")))

@@

Finally, we add a series of informative variable at the course level.
```{r d_course additional info}
d_course <- d_course %>%
  mutate(
    `Missing from ILO File` = as.logical(`Missing from ILO File`),
    Letters = substring(Code, 1, 3),
    Number  = as.numeric(substring(Code, 4, 7)),
    Level   = case_when(between(Number, 1000, 1999) ~ "Introductory",
                        between(Number, 2000, 2999) ~ "Intermediate",
                        between(Number, 3000, 3999) ~ "Advanced"    ))
print(select(d_course, Code, Letters, Number, Level))
```

## Course Catalogue
In order to conduct a topic analysis of course content, we extract the *overview* and *description* of each course from the course catalogues.

### Extracting Overviews
We start by extracting the overviews which are 1 or 2 page long and contain important information of a course. To accomplish this, for each catalogue (loop), we first use `grep` to determine which pages should be excluded from the analysis (`pages_to_exclude`). These pages are typically headers. Excluding them before the analysis allows use to keep the code relatively simple. We also make use the fact that the overviews always start with the code of the course, i.e. it starts with one of the elements of `course_code` and use `grep` to identify the first page of each overview. We use grep a 3rd time to identify the page containing the header *Core Coureses (COR)* and which marks the beginning of the overviews in the catalogue, and we loop from this page to the last page of the catalogue.

@@

### Extracting Descriptions
Since the overviews contain a lot of information that does not interest us, we extract the description section of the overviews. The description section contains a brief description (200-500 words) of the content of the course. To accomplish this, for each overview, we need to identify the line where the description starts and the line where it ends. The description section is always preceeded by a header saying *Description* or *Course Description* (`start_descrip`), making it relatively easy to find the starting line of the decription section with `grep`. The description section is usually followed by the literature section which starts with header saying *Literature*, *Recommended Literature* or similar (see `end_descrip`). Using `grep` together with `end_descrip`, we can again identify the ending line of the description section of almost overviews. A handful overviews do not contain a *Literature* section. For these, the section *Instructional Format* marks the end of the description section.
```{r description}
# Setup
symbol_to_line <- c("\r\n", "\n")
symbol_to_line <- paste(symbol_to_line, collapse = "|")

@@

### Tidy Text Format
We first create two tibbles `d_overview_tidy` and `d_description_tidy` that respectively store the overviews and the descriptions in the tidy text format, with one row per course-year-word.
```{r tidy text data}
d_overview_tidy <- d_description %>%
  select(Code, `Calendar Year`, Overview) %>%
  unnest_tokens(output = word, input = Overview)
print(d_overview_tidy)

@@

### Stemming
We then use the function `hunspell__stem`, which returns valid stems for a given word, to stem the words in the tibbles `d_overview_tidy` and `d_description_tidy`. We first create a function `stem_hunspell` which, given a term, returns its most basic stem (the last one from the list of stem returned by `hunspell__stem`). We then apply `stem_hunspell` on the words of `d_overview_tidy`. Since, `stem_hunspell` is not a vectorized function and the number of words in `d_overview_tidy` is large, we first use distinct on the `word` variable to find the a list containing *once* each term present in the course overviews and apply the function `stem_hunspell` on this list. We save the terms and their stems in the tibble `stems`. Since the course descriptions can be conceptualized as a subset of the course overviews, the tibble `stems` contains all the terms present in the descriptions and can therefore safely used to stem the words in `d_description_tidy`.
```{r function stem_hunspell, cache = TRUE}
stem_hunspell <- function(term) {
  # look up the term in the dictionary
  stems <- hunspell_stem(term)[[1]]

@@

## Course Manual 
We create a dataframe from the Corpus:
```{r}
manuals_td <- tidy(corpus_manuals)
```

@@

```{r}
manuals_tokens <- manuals_td %>%
  select(id,text) %>%
  mutate(Concentration= substring(id, 1, 3), Code= substring(id,1,7))%>%
  unnest_tokens(word, text)
```
###(Optional) Pre-processing Course Manuals- Removing stopwords
Some characters appear with sufficient frequency to be relevant but not often enough accross all courses to be deleted when using tf-idf (see sections bellow), it is therefore convenient to add some stopwords and remove numbers. For this we import stop_words from tidytext, and search all our words with grep for stand alone numbers. We save these words in different dataframes and then exclude with an antijoin.
```{r}

@@

```
##Student Data
###Renaming Variables
First we rename the variable names to make them more comprehensible:
```{r}
d_rename <-d_dirt %>%
  rename(
    ID                     =  "X...Student.Number", 
    Code                   =  "Module..Abbrev..", 
    `Course Title`         =  "Module..Desc..",
    `Program Abbreviation` =  "Program..Abbreviation.", 
    `Program Description`  =  "Program..Description.",               
    Year                   =  "Academic.Year",                      
    Session                =  "Academic.Session",
    `Appraisal Code`       =  "Appraisal.Type",                     
    `Appraisal Title`      =  "Appraisal..Description.",             
    Attempt                =  "Number.rep..attemp",                 
    Grade                  =  "Grade.symbol",                        
    `Appraisal Status`     =  "Appraisal.Status",                   
    `Appraisal Status Description` =  "Appraisal.Status..Desc..",    #is only NA: unique(d4$Appraisal.Status..Desc..)       
    `Booking Reason`       =  "Module.Booking.Reason..Description.",
    `Appraisal Date`       =  "Appraisal.date",                      
    `Attempted Credits`    =  "Attempted.credits",                  
    `Earned Credit`        =  "Earned.Credits",                      
    `Credit Type`          =  "Credit.Type",                        
    `Personnel Number`     =  "Personnel.Number",                  
    `Name (Employee/Applicant)`=  "Name.of.employee.or.applicant",      
    `Individual Work`      =  "Individual.Work",                     
    `Academic Work ID`     =  "Academic.Work.ID",                   
    `Booking Status`       =  "Booking.Status",                      #empty
    `Booking Status Description` =  "Booking.Status..Desc..",             
    `Event Object Id`      =  "Event.Object.Id",                     #empty forwards and including      
    `Business Event Abbreviation` =  "Business.event.abbr.",               
    `Event Description`    =  "Event.Description",                   
    `Start time`           =  "Start.time",                         
    `End time`             =  "End.time",                            
    `Date`                 =  "Date",                               
    `Studyplan`            =  "Studyplan",                           
    `Conditional booking`  =  "Conditiona.booking",                 
    `Wachtlist`            =  "Wachtlist",                           
    `Object ID`            =  "Object.ID",                          
    `Object Name`          =  "Object.name",                         
    `Start Date`           =  "Start.date",                         
    `End Date`             =  "End.Date",                            
    `Number of Weeks`      =  "Number.of.weeks",                      #until this one next has values              
    `Appraiser ID`         =  "Appraiser.ID",                        
    `Appraiser Name`       =  "Appraiser.Name"
